---
title: "Is Chelsea boring?"
author: "Mihai Sirbu"
date: "Monday, September 07, 2015"
output: pdf_document
---

```{r, echo=FALSE, include=FALSE}
library(plyr)
library(dplyr)
library(stringr)
library(ggplot2)
```

#Intro 

Jose Mourinho returned to Chelsea on June 3rd 2013. As soon as he did, many 
premier league fans (including myself) braced for a return of "boring Chelsea."
In fact, the idea that Chelsea like to "park the bus" had been a complaint even
before Mourinho returned. Which made me want to know: is this even true?

(Full disclaimer: I'm a Liverpool fan and I would like nothing more than
confirming this belief. Keep this in mind when I discuss the results.)

To investigate this question I used the english premier league results dataset
from James P. Curley (2015). engsoccerdata: English Soccer Data 1871-2015.
(You can check out his github at: https://github.com/jalapic/engsoccerdata).

Because I started watching the premier league starting from the 2009 season, I
decided to investigate "boringness" (defined below) from the 2009-2010 to the
2014-2015 season. I also decided to use the teams that have been often considered
the top 7 teams in the premier league: Liverpool, Manchester United, Chelsea,
Arsenal, Manchester City, Tottenham, and Everton. 

To be able to access the data do the following:
```{r}
#Commented out because already done.
#library(devtools)
#install_github('jalapic/engsoccerdata', username = "jalapic")
library(engsoccerdata)
```

Then you can load the data as follows:
```{r}
data(engsoccerdata2)
engsoccer_tbl <- as.tbl(engsoccerdata2)
```

Let's get a sense of the data and output the first ten rows
```{r}
engsoccer_tbl
str(engsoccer_tbl)
```

#Prep

Before we get to the analysis, we need to prepare the necessary components and
table

```{r}
#The teams we wish to examine
teams <- c("Arsenal", "Chelsea", "Everton", "Liverpool", "Manchester City", 
           "Manchester United", "Tottenham Hotspur")

#We want to start on the 2009 season and go until the most recent season
rel_tbl <- engsoccer_tbl %>% filter(Season > 2008, division == 1) %>% 
  filter(home %in% teams | visitor %in% teams)

#rel_tbl$home <- str_replace_all(rel_tbl$home, " ", "_")
#rel_tbl$visitor <- str_replace_all(rel_tbl$visitor, " ", "_")

#team_result <- lapply(teams, function(team){return(matrix(0, nrow=10,ncol=10))})
#names(team_result) <- str_replace_all(teams," ", "_")
#teams <- str_replace_all(teams, " ", "_")
```

Now what do I mean by boring? I mean "boring games" where I define as scores 
that are contained as 0-0, 1-0, or 0-1. To figure this out we need to figure 
out what percentage of games end in each score line up to a maximum of9 goals,
the game between Tottenham and Wigan
```{r}
max(max(rel_tbl$hgoal),max(rel_tbl$vgoal))


```