---
title: "Is Chelsea boring?"
author: "Mihai Sirbu"
date: "Monday, September 07, 2015"
output: pdf_document
---

```{r, echo=FALSE, include=FALSE}
library(plyr)
library(dplyr)
library(stringr)
library(ggplot2)
```

#Intro 

Jose Mourinho returned to Chelsea on June 3rd 2013. As soon as he did, many 
premier league fans (including myself) braced for a return of "boring Chelsea."
In fact, the idea that Chelsea like to "park the bus" had been a complaint even
before Mourinho returned. Which made me want to know: is this even true?

(Full disclaimer: I'm a Liverpool fan and I would like nothing more than
confirming this belief. Keep this in mind when I discuss the results.)

To investigate this question I used the english premier league results dataset
from James P. Curley (2015). engsoccerdata: English Soccer Data 1871-2015.
(You can check out his github at: https://github.com/jalapic/engsoccerdata).

Because I started watching the premier league starting from the 2009 season, I
decided to investigate "boringness" (defined below) from the 2009-2010 to the
2014-2015 season. I also decided to use the teams that have been often considered
the top 7 teams in the premier league: Liverpool, Manchester United, Chelsea,
Arsenal, Manchester City, Tottenham, and Everton. 

To be able to access the data do the following:
```{r}
#Commented out because already done.
#library(devtools)
#install_github('jalapic/engsoccerdata', username = "jalapic")
library(engsoccerdata)
```

Then you can load the data as follows:
```{r}
data(engsoccerdata2)
engsoccer_tbl <- as.tbl(engsoccerdata2)
```

Let's get a sense of the data and output the first ten rows
```{r}
engsoccer_tbl
str(engsoccer_tbl)
```

#Prep

Before we get to the analysis, we need to prepare the necessary components and
table

```{r}
#The teams we wish to examine
teams <- c("Arsenal", "Chelsea", "Everton", "Liverpool", "Manchester City", 
           "Manchester United", "Tottenham Hotspur")

#We want to start on the 2009 season and go until the most recent season
rel_tbl <- engsoccer_tbl %>% filter(Season > 2008, division == 1) %>% 
  filter(home %in% teams | visitor %in% teams)

#rel_tbl$home <- str_replace_all(rel_tbl$home, " ", "_")
#rel_tbl$visitor <- str_replace_all(rel_tbl$visitor, " ", "_")

#team_result <- lapply(teams, function(team){return(matrix(0, nrow=10,ncol=10))})
#names(team_result) <- str_replace_all(teams," ", "_")
#teams <- str_replace_all(teams, " ", "_")
```

Now what do I mean by boring? I mean "boring games" where I define as scores 
that are contained as 0-0, 1-0, or 0-1. To figure this out we need to figure 
out what percentage of games end in each score line up to a maximum of 9 goals,
the game between Tottenham and Wigan
```{r}
max_goals <- max(max(rel_tbl$hgoal),max(rel_tbl$vgoal))

rel_tbl %>% filter(hgoal == max_goals | vgoal == max_goals)

```

To calculate the necessary percentages we are going to use the following 
function:

```{r}
calculate_team_percent <- function(team, data){
  total_count <- data %>% filter(home == team | visitor == team) %>%
  mutate(Team = team, Team_Goals_Scored = ifelse(home == team, hgoal, vgoal), 
         Opp_Goals_Scored = ifelse(home == team, vgoal, hgoal)) %>%
         group_by(Team, Team_Goals_Scored, Opp_Goals_Scored) %>% 
         tally() %>% 
         ungroup() %>%
         mutate(Percent = n/sum(n)* 100)
  
  all_fixtures <- as.tbl(cbind(Team=team, expand.grid(Team_Goals_Scored=seq(0,9),
                                                Opp_Goals_Scored=seq(0,9))))
                         
  merged_tbl <- as.tbl(merge(all_fixtures, total_count, all.x=TRUE, 
                      by=c("Team","Team_Goals_Scored","Opp_Goals_Scored")))
  
  merged_tbl <- mutate(merged_tbl, n = ifelse(is.na(n), 0, n), 
                      Percent = ifelse(is.na(Percent), 0, Percent))
  return(merged_tbl)
}

```

Let me stop and explain what the function above is doing. Essentially it's 
creating two separate tables and joining them together. The first table (total
count) tallies up all of the different score combinations across the last six 
seaons. So it tells you, for example, the percentage of games that end 0-0. 
Unfortunately, it does not include scores that DO NOT occur such as 7-7. 
However for each team we want to create a heatmap and we want to show that 
particular scoreline to be 0. Which is where the second table comes in. 
In all_fixtures we are creating a table with all possible score lines up to
9-9. From there we outer join the two tables (thus giving us a table with all
fixtures included) and replace all NAs (because fixtures not present in 
total_count are given a value of NA) in the n and Percent column with 0.

We can get the results for all teams by doing the following: 
```{r}
all_teams <- as.tbl(do.call("rbind", lapply(teams, calculate_team_percent,data=rel_tbl)))

```


#Analysis 

Now that we have the desired data frame, it's time we created some heatmaps. 
We're going to use this function below to create a heatmaps. 

```{r}
create_heatmap <- function(team, data){
  
  team_data <- data %>% filter(Team == team)
                               
  ggplot(team_data, aes(x=Opp_Goals_Scored,y=Team_Goals_Scored)) + 
  geom_tile(aes(fill = Percent),colour = "white") + 
geom_text(aes(fill = Percent, label = round(Percent, 2)))+
  scale_fill_gradient(low = "white", high = "steelblue")+
  scale_x_continuous(breaks=0:9, expand=c(0,0))+
  scale_y_continuous(breaks=0:9, expand=c(0,0))+
  ggtitle(str_c("Distribution of ", team, "\nFixture Results (2009-2015)"))+
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))+
  labs(x="# goals scored by Opponent")+
  labs(y=str_c("# goals scored by ", team))
}

```

Here's the heatmap distribution for Liverpool

```{r}
create_heatmap("Liverpool", all_teams)
```

We can easily create the 6 other heatmaps but it's not easy to compare across
7 plots. Instead we're going to filter the all_teams dataframe to easily compare
across teams. 

\pagebreak

It's time to now address some of the questions we hope to address.

##Who are the boring teams?
```{r}
all_teams %>% group_by(Team) %>% filter(Team_Goals_Scored <= 1, Opp_Goals_Scored <= 1) %>%
  summarise(Percent_Boring = sum(Percent)) %>% arrange(desc(Percent_Boring))
```

As you can see here, Everton actually tops the list at a little over 38.5% The other 
6 teams are all in roughly the same range, with a low of ~29% and a high of ~33%.
From this evidence alone we cannot say that Chelsea is "boring."

##Does Chelsea dominate in the 1-0 category?
```{r}
all_teams %>% group_by(Team) %>% filter(Team_Goals_Scored == 1, Opp_Goals_Scored == 0) %>%
  summarise(Grind_Wins = sum(Percent)) %>% arrange(desc(Grind_Wins))
```

Another complaint is that Chelsea are the experts in grinding out "1-0" wins.
From this evidence we can't say that Chelsea are any better than any other team.

###Or the 0-0 category?
```{r}
all_teams %>% group_by(Team) %>% filter(Team_Goals_Scored == 0, Opp_Goals_Scored == 0) %>%
  summarise(No_Score = sum(Percent)) %>% arrange(desc(No_Score))
```

Maybe Chelsea are specialists in goalless draws. Even in this case, that is false.
In fact, my beloved Liverpool tops this category (but only ~1% more than the
majority of the other teams) while the team I dread the most has the lowest
percentage of games that end in goalless draws across these 6 seasons. 